Time Series Analysis of Chaotic Dynamical Systems with Recurrent Neural Networks

### 2019-04-16 doc/paper/paper.tex

One of the defining characteristics of chaos is that the error from
perturbations grows exponentially \cite{oestreicher2007history}. At the time
of Lorenz's paper, unstable error was not well understood, and this led to
Lorenz's surprising discovery \cite{oestreicher2007history}. Lorenz reported
that a single iteration in calculating the solution to the dynamical system
took approximately one second \cite{lorenz1963deterministic}. In his paper,
he shows one of his calculations having about 3000 iterations, which must
have taken around 50 minutes. Given the amount of time it takes to perform
the calculation, it's tempting to continue the calculation from the output of
the previous calculation. However, the story of the discover is that while
the computer used 6 digit accuracy, it output only 3 digits, so when the
simulation was continued by Lorenz with the less accurate measurements, he
found the results to be very different \cite{oestreicher2007history}.

@@

In order to find the Lyapunov time, we only need to solve for the largest
LCE. To find this, we will use the following method detailed in
\cite{viswanath1998lyapunov}.

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|l|}
         % TODO: \lambda_1?
         % TODO: calculated values
        \hline
        $\sigma$ & $\rho$ & $\beta$ & actual $\lambda_1$ & calculated $\lambda_1$ & relative error of calculated $\lambda_1$ \\
        \hline \hline
        16 & 45.92 & 4 & 1.50255 & & \\ % DATA POINTS
        16 & 40 & 4 & 1.37446 & & \\
        10 & 28 & 8/3 & 0.90566 & & \\
        \hline
    \end{tabular}
    \caption{
        Lorenz Parameters and Largest Lyapunov Exponent
        \cite{viswanath1998lyapunov}
    }
    \label{table:lorenz_params}
\end{table}

@@

While it is possible to use a symplectic integrator, it is not necessary.
Under the right initial conditions, the third body will diverge regardless.
The simple divergent spiral isn't chaotic, so instead orbital and collision
less initial conditions will be used. Without divergence and without
collisions, the symplectic integrator is unnecessary.

@@

The output layer is connected to the current input layer and current 
internal layer. The weight for these connections will be denoted $\Wout$ 
(we only have one matrix since the connected layers are concatenated into a
single vector, per Jaeger's notation \cite{jaeger2002tutorial}). The 
activation function for these connections is linear. Although, the 
recommended choice
is $\tanh$, the output layer doesn't necessarily lie in the range of 
$\tanh$, so it can't be used without some form of normalization.

@@

\begin{align}
    \xv(t_{i}) &= \tanh \left(
            \Win \uv(t_i)
            + \W \xv(t_{i-1})
            + \Wback \yv(t_{i-1})
        \right) \\
    \yv(t_i) &= \Wout (\uv(t_i), \xv(t_i))_{\text{concat}}.
\end{align}

@@

Third, the training values are fed into the model to sample the network.
We can store the input unit values, the internal state values, and the 
desired output values as matrices $M$ and $T$. $M$ will store the input 
unit values and internal state value pairs
$(\uv(t_{T_0}), \xv(t_{T_0})), \dots, (\uv(t_T), \xv(t_T))$
in its rows. $T$ will store the output unit values
$\dv(t_{T_0}), \dots, \dv(t_T)$.

@@

\subsection{Method}
\subsection{Results}

>>>>

One of the defining characteristics of chaos is that the error from
perturbations grows exponentially \cite{oestreicher2007history}. At the time
of Lorenz's paper, unstable error was not well understood, and this led to
Lorenz's surprising discovery \cite{oestreicher2007history}. Lorenz reported
that a single iteration in calculating the solution to the dynamical system
took approximately one second \cite{lorenz1963deterministic}. In his paper,
he shows one of his calculations having about 3000 iterations, which must
have taken around 50 minutes. Given the amount of time it takes to perform
the calculation, it's tempting to continue the calculation from the output of
the previous calculation. However, the story of the discover is that while
the computer used 6 digit accuracy, it only output 3 digits, so when the
simulation was continued by Lorenz with the less accurate measurements, he
found the results to be very different \cite{oestreicher2007history}.

@@

\label{sec:calc_lya}
In order to find the Lyapunov time, we only need to solve for the largest
LCE. To find this, we will use the following method detailed in
\cite{viswanath1998lyapunov}.

@@

\begin{table}[H]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|l|}
        \hline
        $\sigma$ & $\rho$ & $\beta$ & actual $\lambda_1$ & calculated $\lambda_1$ & relative error of calculated $\lambda_1$ \\
        \hline \hline
        4 & 45.92 & 16 & 1.50255 & 1.49869 & -2.5694 $\times 10^{-3}$ \\ % DATA POINTS
        4 & 40 & 16 & 1.37446 & 1.37297 & -1.0876 $\times 10^{-3}$ \\
        8/3 & 28 & 10 & 0.90566 & 0.91188 & 6.8694 $\times 10^{-3}$ \\
        \hline
    \end{tabular}
    \caption{
        Lorenz Parameters and Largest Lyapunov Exponent
        \cite{viswanath1998lyapunov}
    }
    \label{table:lorenz_params}
\end{table}

@@

While it is possible to use a symplectic integrator, it is not necessary.
Under the right initial conditions, the third body can diverge regardless
(fast initial velocity away from the bodies). The simple divergent spiral
isn't chaotic, so instead orbital and collision less initial conditions
will be used. Without divergence and without collisions, the symplectic
integrator is unnecessary.

@@

The output layer is connected to the current input layer and current 
internal layer. Additionally, a bias input is connected to this layer.
The weight for these connections will be denoted $\Wout$ 
(we only have one matrix since the connected layers are concatenated into a
single vector, per Jaeger's notation \cite{jaeger2002tutorial}). The 
activation function for these connections is linear. Although, the 
recommended choice
is $\tanh$, the output layer doesn't necessarily lie in the range of 
$\tanh$, so it can't be used without some form of normalization.

@@

\begin{align}
    \xv(t_{i}) &= \tanh \left(
            \Win \uv(t_i)
            + \W \xv(t_{i-1})
            + \Wback \yv(t_{i-1})
        \right) \\
    \yv(t_i) &= \Wout (\uv(t_i), 1, \xv(t_i))_{\text{concat}}.
\end{align}

@@

Third, the training values are fed into the model to sample the network.
We can store the input unit values, the internal state values, and the 
desired output values as matrices $M$ and $T$. $M$ will store the input 
unit values and internal state value pairs
$(\uv(t_{T_0}), 1, \xv(t_{T_0})), \dots, (\uv(t_T), 1, \xv(t_T))$
in its rows. $T$ will store the output unit values
$\dv(t_{T_0}), \dots, \dv(t_T)$.

@@

\subsection{Method}

First, data from the models is collected. Using the process described
above (Section \ref{sec:calc_lya}), the Largest Lyapunov Exponent is
calculated, $\lambda_1$. Then, the Lyapunov Time is calculated by
$\frac{1}{\lambda_1}$. We will then integrate the model with an algorithm
called RK4(5) over the interval
$[0, \frac{50}{\lambda_1}]$ with time steps of $\frac{1}{100 \lambda_1}$. 
This will result in 5,000 data points.

Then, the data is partitioned into a training set, a validation set, and 
a testing set. I use a 90\% split, so testing is the last 10\% of the
data. The training and validation set are the first 90\%. Of the 90\%,
training is the first 90\%, and validation is the final 10\%. In total,
it's a 81\%-9\%-10\% split.

The training set is used to train the model. The validation
set is used to tune the parameters of the model ($\alpha, N, T_0$). The
testing set is used to evaluate the performance.

The first data point of the training set is $t_0 = 0$. The parameter
$T_0$ marks where the model begins training on the data (see Section
\ref{sec:etrain}). Using slightly different notation to account for the
extra validation set, we will denote the first validation time point by
$t = T$. We will denote the first testing time point by $t = T_f$

Performance of the model is calculated by RMSE for a given time interval
$[t_0, t_f]$:

\begin{align}
    \text{RMSE} = \sqrt{\sum^{t_f}_{t=t_0} (\dv(t) - \yv(t))^2}
\end{align}

where $(\cdot)^2$ is the dot product of the error with itself,
$\dv(t)$ is the true value at time $t$,
and $\yv(t)$ is the model's approximation at time $t$.

First, we generate a list of parameter combinations to try. Usually, this
is the Cartesian product of typical parameter values for the model. 
For each combination, we train several models on the training data. Then,
we predict the validation data to measure the general performance of the
model under those parameters.

$\text{RMSE}_\text{train}$ is calculated on the interval $[T_0, T)$,
and the prediction uses teacher-forced values on $[0, T_0)$.
$\text{RMSE}_\text{validation}$ is calculated on the interval $[T, T_f)$,
and the prediction uses teacher-forced values on $[0, T)$.

Then, the models which performed best on the validation set should also
perform well on the testing set. The models are retrained on the testing
and validation set, and the error is calculated:
$\text{RMSE}_\text{full train}$ is calculated on the interval $[T, T_f)$,
and the prediction uses teacher-forced values on $[0, T)$.

Finally, the error for the testing set is calculated. 
The performance on the testing set is 
an unbias representation of how well the ESN can fit the model.

However, the $\text{RMSE}_\text{test}$ is calculated slightly different
since we want to know how far the model can predict before deviating too
far. The $\text{RMSE}_\text{test}(t_f)$ will be calculated on the interval
$[T_f, t_f)$ and the prediction uses teacher-forced values on $[0, T_f)$.
Lastly, $\text{RMSE}_\text{test}$ will notate the RMSE for the entire test
set.

We will look at how far past $T_f$ the ESN model's
$\text{RMSE}_\text{test}$ grows too large.

\subsection{Results}
\subsubsection{Lorenz}

One of the issues run into while forecasting with the Lorenz is that the
continuous nature of the equation led to the ESN just outputing the
previous value. For initial values far from the center of the Lorenz
Butterfly, it would tend towards the center and stop when it does arrive
there. Most models trained poorly and gave $\text{RMSE}_\text{test}$
values of 16 or well above 10,000, implying the model diverged.

To fix this issue, we instead have the ESN train on differences between
time points. This reduces our data set to 4,999 values, and it creates
a new shape of the data. The Lorenz Butterfly had a similar shape, two rings, but they intersected differently (See figure \ref{fig:lorenz_diff}).

\begin{figure}[H]
    \centering
    \includegraphics[width=.5\linewidth]{images/lorenz/test.png}
    \caption{Largest Lyapunov Characteristic Exponent vs Inner Theta Initial Condition}
    \label{fig:lorenz_diff}
\end{figure}

\subsubsection{Double Pendulum}
\subsubsection{Restricted Circular 3 Body}
